Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
add cross-validation

library(tidyverse)
library(arrow)
library(knitr)
library(usmap)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(rstanarm)
library(fixest)  
```

@@

In this study, we utilized several R packages to streamline data
manipulation, modeling, and visualization. The `tidyverse` package
@tidyverse was central to our workflow, providing tools for data
wrangling and analysis, while `arrow` @arrow efficiently managed parquet
files for larger datasets. We used `modelsummary` package @modelsummary
to generate clean and interpretable summaries of our regression
models.We used `knitr` @knitr and `kableExtra` @kableExtra to generate
clean tables and reproducible reports, and `usmap` for geographical
visualizations. We use `here` @here package simplifies the management of
file paths, ensuring a reproducible workflow by creating a consistent
point of reference for file locations, regardless of the operating
system or working directory. Data cleaning was facilitated by `janitor`
@janitor, while `lubridate` @lubridate handled date-time variables.
Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
regression modeling was conducted using `rstanarm` @rstanarm. We use `fixest` @fixest to conduct the event study. Together, these packages supported an efficient, reproducible workflow for
analyzing and presenting our findings.

@@

The results from the two Bayesian models are summarized in
@tbl-modelresults. The interaction term between after_injunction and
abortion_illegal is positive and statistically significant (0.111 in
Model 1 and 0.280 in Model 2), indicating a change in infant mortality
rates associated with the combination of abortion bans and the
post-injunction period.

For the independent effects, states where abortion is illegal show
higher infant mortality rates, with coefficients of 3.365 in Model 1 and
1.400 in Model 2. The post-injunction period alone is associated with
coefficients of -0.525 in Model 1 and -0.940 in Model 2, indicating a
decrease in infant mortality during this time.

Demographic variables also exhibit some associations. Maternal age is
negatively associated with infant mortality, with coefficients for age
groups 25–29 and 30–34 of -5.73 and -6.15 in Model 2, relative to group 15-19 respectively. The
the coefficient for Black mothers in Model 2 is 6.66 relative to Asia monthers, reflecting a higher
infant mortality compared to other racial groups.

@@

The analysis reveals notable patterns in infant mortality rates influenced by policy and demographic factors. For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in Model 1 and 0.947 in Model 2, compared to the pre-injunction period. In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in Model 1 and 1.401 in Model 2, relative to states where abortion remained legal.

The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in Model 1 and 0.280 in Model 2. These effects highlight the compounded impact of restrictive abortion policies on infant health outcomes.

Demographic factors also play a significant role. Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in Model 2, respectively. Regarding maternal race, Black or African American mothers have a significantly higher infant mortality rate, with an increase of 6.604 deaths per 1,000 live births compared to Asian mothers (reference level). In contrast, White mothers exhibit a lower infant mortality rate, with a decrease of 2.562 deaths per 1,000 live births compared to Asian mothers.

@@

In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model. It represents posterior predictive checks for Model 1 and Model 2, illustrating how well each model captures the observed data. The posterior predictive checks for Model 1 (policy-focused) and Model 2 (policy and demographic factors) illustrate model fit. In @fig-ppcheckandposteriorvsprior-1, the observed ($y$) and replicated ($y_{rep}$) distributions align moderately, reflecting general trends but limited precision. @fig-ppcheckandposteriorvsprior-2 shows improved alignment, indicating that adding demographic factors enhances predictive accuracy and better captures the complexities of infant mortality rates.

In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior. The posterior vs. prior plots for Model 1 and Model 2 illustrate the role of data in informing the posterior estimates of the model parameters. In Model 1 @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power. Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables. The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.

@@

plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```
Credibility intervals are the Bayesian equivalent of confidence intervals, offering an understanding of the range of probable values for a parameter @tellingstories. Using Bayesian estimation, we derive a posterior distribution for each model coefficient. @fig-area-1 presents the 95% credibility intervals for coefficients in Model 1, while @fig-area-2 illustrates the same for Model 2. These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.

@@

The trace plot in @fig-stanareyouokay-1 shows the sampling chains for model 1, indicating good mixing and convergence as the chains overlap and appear stable throughout. This suggests that the model has achieved convergence during sampling. The Rhat plot in @fig-stanareyouokay-2 demonstrates that all Rhat values are below 1.05, further confirming convergence and indicating that the model’s posterior samples are reliable. @fig-stanareyouokay-3 represents the trace plots for Model 2, showing the posterior samples across four chains. The plots indicate good mixing, as the chains overlap well and stabilize around the same region, suggesting convergence.
@fig-stanareyouokay-4 is the Rhat diagnostic plot for Model 2. All Rhat values are close to 1.00, confirming that the Markov chains have converged, and the posterior estimates are reliable. This supports the validity of Model 2’s Bayesian estimation process.

>>>>

library(tidyverse)
library(arrow)
library(knitr)
library(usmap)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(rstanarm)
library(fixest)  
library(caret)
```

@@

In this study, we utilized several R packages to streamline data
manipulation, modeling, and visualization. The `tidyverse` package
@tidyverse was central to our workflow, providing tools for data
wrangling and analysis, while `arrow` @arrow efficiently managed parquet
files for larger datasets. We used `modelsummary` package @modelsummary
to generate clean and interpretable summaries of our regression
models.We used `knitr` @knitr and `kableExtra` @kableExtra to generate
clean tables and reproducible reports, and `usmap` for geographical
visualizations. We use `here` @here package simplifies the management of
file paths, ensuring a reproducible workflow by creating a consistent
point of reference for file locations, regardless of the operating
system or working directory. Data cleaning was facilitated by `janitor`
@janitor, while `lubridate` @lubridate handled date-time variables.
Visualization tasks were enhanced with `ggplot2` @ggplot2, and Bayesian
regression modeling was conducted using `rstanarm` @rstanarm. We use `fixest` @fixest to conduct the event study. We use `caret` @caret to conduct the cross-validation of model. Together, these packages supported an efficient, reproducible workflow for
analyzing and presenting our findings.

@@

The results from the two Bayesian models are summarized in
@tbl-modelresults. The interaction term between after_injunction and
abortion_illegal is positive and statistically significant (0.111 in
first model and 0.280 in second model), indicating a change in infant mortality
rates associated with the combination of abortion bans and the
post-injunction period.

For the independent effects, states where abortion is illegal show
higher infant mortality rates, with coefficients of 3.365 in first model and
1.400 in second model. The post-injunction period alone is associated with
coefficients of -0.525 in first model and -0.940 in second model, indicating a
decrease in infant mortality during this time.

Demographic variables also exhibit some associations. Maternal age is
negatively associated with infant mortality, with coefficients for age
groups 25–29 and 30–34 of -5.73 and -6.15 in second model, relative to group 15-19 respectively. The
the coefficient for Black mothers in second model is 6.66 relative to Asia monthers, reflecting a higher
infant mortality compared to other racial groups.

@@

The analysis reveals notable patterns in infant mortality rates influenced by policy and demographic factors. For states where abortion remained legal, the post-injunction period (after_injunction1) is associated with reductions in infant mortality rates, with a decrease of 0.525 deaths per 1,000 live births in first model and 0.947 in second model, compared to the pre-injunction period. In contrast, states where abortion became illegal (abortion_illegal1) experienced significantly higher infant mortality rates, with increases of 3.365 deaths in first model and 1.401 in second model, relative to states where abortion remained legal.

The Difference-in-Differences (DID) interaction term (after_injunction1 × abortion_illegal1) indicates that abortion bans during the post-injunction period resulted in an additional increase of 0.111 deaths per 1,000 live births in first model and 0.280 in second model. These effects highlight the compounded impact of restrictive abortion policies on infant health outcomes.

Demographic factors also play a significant role. Compared to mothers aged 15–19 years (reference level), older maternal age groups (20–24, 25–29, and 30–34 years) are associated with progressively lower infant mortality rates, with reductions of 3.571, 5.732, and 6.155 deaths per 1,000 live births in second model, respectively. Regarding maternal race, Black or African American mothers have a significantly higher infant mortality rate, with an increase of 6.604 deaths per 1,000 live births compared to Asian mothers (reference level). In contrast, White mothers exhibit a lower infant mortality rate, with a decrease of 2.562 deaths per 1,000 live births compared to Asian mothers.

@@

In @fig-ppcheckandposteriorvsprior-1 and @fig-ppcheckandposteriorvsprior-2 we implement a posterior predictive check for the first model and second model. It represents posterior predictive checks for first model and second model, illustrating how well each model captures the observed data. The posterior predictive checks for first model (policy-focused) and second model (policy and demographic factors) illustrate model fit. In @fig-ppcheckandposteriorvsprior-1, the observed ($y$) and replicated ($y_{rep}$) distributions align moderately, reflecting general trends but limited precision. @fig-ppcheckandposteriorvsprior-2 shows improved alignment, indicating that adding demographic factors enhances predictive accuracy and better captures the complexities of infant mortality rates.

In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior. The posterior vs. prior plots for first model and second model illustrate the role of data in informing the posterior estimates of the model parameters. In first model @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power. second model @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables. The comparison indicates that second model provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.

@@

plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```
Credibility intervals are the Bayesian equivalent of confidence intervals, offering an understanding of the range of probable values for a parameter @tellingstories. Using Bayesian estimation, we derive a posterior distribution for each model coefficient. @fig-area-1 presents the 95% credibility intervals for coefficients in first model, while @fig-area-2 illustrates the same for second model. These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.

## Cross Validation

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
set.seed(123)

# Split the data into training and testing sets (80/20 split)
train_index <- createDataPartition(analysis_data$death_rate, p = 0.8, list = FALSE)
train_data <- analysis_data[train_index, ]
test_data <- analysis_data[-train_index, ]

# Fit a Bayesian model on the training set
model1 <- stan_glmer(
  death_rate ~ after_injunction * abortion_illegal+ (1 | state),
  data = train_data,
  family = gaussian(),
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 2.5),
  prior_aux = exponential(1),
  iter = 2000, chains = 4, seed = 123
)
# Make predictions on the test set
predictions <- posterior_predict(model1, newdata = test_data)
predicted_means <- rowMeans(predictions)  # Average over posterior samples

# Calculate RMSE manually
errors <- test_data$death_rate - predicted_means
rmse_value <- sqrt(mean(errors^2))

# Output the RMSE
print(paste("The RMSE is:", round(rmse_value, 2)))
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
# Fit a Bayesian model on the training set
model2 <- stan_glmer(
  death_rate ~ after_injunction * abortion_illegal + year_of_death + age_of_mother + mothers_single_race + (1 | state),
  data = train_data,
  family = gaussian(),
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 2.5),
  prior_aux = exponential(1),
  iter = 2000, chains = 4, seed = 123
)
# Make predictions on the test set
predictions <- posterior_predict(model2, newdata = test_data)
predicted_means <- rowMeans(predictions)  # Average over posterior samples

# Calculate RMSE manually
errors <- test_data$death_rate - predicted_means
rmse_value <- sqrt(mean(errors^2))

# Output the RMSE
print(paste("The RMSE is:", round(rmse_value, 2)))
```
The RMSE values from cross-validation indicate that first model and second model perform similarly in predicting infant mortality rates, with RMSE values of 5.07 and 5.06, respectively. This similarity suggests that adding demographic factors such as maternal age and race in second model has only a marginal impact on predictive accuracy compared to first model, which primarily relies on policy-related variables like abortion legality and timing of the injunction. An average error of approximately 5 deaths per 1,000 live births indicates the magnitude of prediction errors, which may or may not be acceptable depending on the variability in the dataset. While second model slightly reduces the error, its added complexity may not justify the trade-off if simplicity is a priority. However, if a nuanced understanding of demographic influences is essential, second model provides a modest improvement in accuracy while offering more detailed insights into maternal and infant health disparities. The cross-validation process reinforces the reliability of these models by confirming their performance across multiple data splits, reducing the risk of overfitting.

@@

The trace plot in @fig-stanareyouokay-1 shows the sampling chains for first model, indicating good mixing and convergence as the chains overlap and appear stable throughout. This suggests that the model has achieved convergence during sampling. The Rhat plot in @fig-stanareyouokay-2 demonstrates that all Rhat values are below 1.05, further confirming convergence and indicating that the model’s posterior samples are reliable. @fig-stanareyouokay-3 represents the trace plots for second model, showing the posterior samples across four chains. The plots indicate good mixing, as the chains overlap well and stabilize around the same region, suggesting convergence.
@fig-stanareyouokay-4 is the Rhat diagnostic plot for second model. All Rhat values are close to 1.00, confirming that the Markov chains have converged, and the posterior estimates are reliable. This supports the validity of second model’s Bayesian estimation process.
